Assessing linkages between watershed nutrient loading and estuary water quality in Lavaca Bay, Texas

### 2023-02-11 Schramm-Manuscript-2023/Schramm-Manuscript-2023.tex
edits. Add IT approach to methods, add lagged responses in discussion.

@@

Based on criteria provided by \citet{moriasiHydrologicWaterQuality2015},
GAMs ranged from ``satisfactory'' to ``very good'' using median NSE,
r\textsuperscript{2} and PBIAS calculated using repeated 5-fold cross
validation on predicted and measured nutrient loads. NO\textsubscript{3}
GAM models had median NSE values of 0.34 and 0.48 at USGS-08164000 and
USGS-08164390 respectively. Median r\textsuperscript{2} values were 0.70
(USGS-08164000) and 0.87 (USGS-08164525) and PBIAS values were 2.00
(USGS-08164000) and 10.90 (USGS-08164525) for NO\textsubscript{3} loads.
Median NSE values were 0.80 (USGS-08164000) and 0.91 (USGS-08164525) for
TP loads. Median r\textsuperscript{2} values for TP loads were 0.93
(USGS-08164000) and 0.99 (USGS-08164525) and median PBIAS values were
-7.20 (USGS-08164000) and -3.30 (USGS-08164525). Density plots of
metrics show similar distribution of values between sites for the same
parameter, with the exception r\textsuperscript{2} values for
NO\textsubscript{3} loads where USGS-08164000 showed a much larger
variance in values compared to USGS-08164525 (Figure \ref{fig:fig2}). TP
GAMS had higher average NSE and r\textsuperscript{2} values and lower
variance in metric values compared to NO\textsubscript{3}.

@@

Annual NO\textsubscript{3} and TP loads show considerable variation,
generally following patterns in discharge (Figure \ref{fig:fig3}, Figure
\ref{fig:fig4}). Flow-normalized TP loads at both sites and
flow-normalized loads at USGS-08164000 indicate watershed based loads
have not changed much over time when accounting for changes in
streamflow (Figure \ref{fig:fig3}). Flow-normalized loads at
USGS-08164000 show small variation over time with some decreases in
NO\textsubscript{3} loads since 2013. Aggregated across both sites, the
mean annual NO\textsubscript{3} load 2005 through 2020 was 205,405 kg
(126,867 kg - 341,569 kg, 90\% CI). Annual NO\textsubscript{3} loads
ranged from 12,574 kg in 2011 to 794,510 kg in 2007.

Total annual TP loads ranged from 7,839 kg in 2011 to 595,075 kg in
2007. Mean annual TP loading from 2005 through 2020 was 182,673 kg
(152,227 kg - 219,310 kg, 90\% CI). On average, USGS-08164525 accounted
for 68\% of NO\textsubscript{3} loads and 59\% of TP loads from 2005
through 2020. However, during periods of extreme drought the Lavaca
River (USGS-08164000) became the primary source of nutrient loading in
the watershed with the Navidad River only accounting for 15\% and 25\%
of NO\textsubscript{3} and TP loads in 2011 (Figure \ref{fig:fig4}).

@@

Other studies have published estimates of TP loading estimates for the
Lavaca River (USGS-08164000) over various time periods (Table
\ref{tab:tpcomp}). Converted to average annual yield, the estimates of
TP loading in this study are within the ranges of previous studies
\citep{dunnTrendsNutrientInflows1996, rebich_sources_2011, omaniEstimationSedimentNutrient2014, wise_spatially_2019},
with the TP estimates in \citet{dunnTrendsNutrientInflows1996} being
notably lower. Given that none of the studies identify substantially
sized trends in TP, it is is possible that the period used in
\citet{dunnTrendsNutrientInflows1996} was drier on average than the
other studies. The SPARROW models used in \citet{rebich_sources_2011}
and \citet{wise_spatially_2019} utilize a version of LOADEST in the
underlying load estimation procedure, so a difference due to methodology
alone is unlikely.

@@

Cross-validation of the GAM loading models indicated that GAMs performed
well on average at predicting daily nutrient loading values. The
variance in scores was very high indicating subsets of values were
problematic at characterizing functional relationships between nutrients
and predictors. All of the water quality data for the two river sites in
the TCEQ database were ambient water quality data, collected to be
representative of typical conditions. This results in few data collected
at the highest portions of the flow duration curve under which the
majority of loadings take place. Supplementary flow-biased monitoring
targeting storm- or high-flow conditions is recommended here to improve
the precision of GAM predictions
\citep{horowitzEvaluationSedimentRating2003, snelderEstimationCatchmentNutrient2017}.

The non-linear temporal water quality trends identified using GAMs
differed slightly from the trends identified by
\citet{bugica_water_2020}. This is not unexpected due to the different
time periods, different methodology, and generally small slopes
identified for most of the significant water quality parameters in prior
work. The trend in DO and cholorophyll-\emph{a} concentrations are
stable in comparison to other Texas estuaries that might be facing
larger demands for freshwater diversions, higher population growth, and
more intense agricultural production
\citep{wetzWaterQualityDynamics2016, bugica_water_2020}. The trend of
increasing NO\textsubscript{\emph{x}} concentration at the upper-bay
TCEQ-13563 site and recent increases in TKN concentration at the mid-bay
TCEQ-13383 site are concerning due to the nitrogen limitation identified
in many Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the relatively low ambient concentrations observed in Lavaca Bay.

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{x}}, TKN, and TP are suggestive of nonpoint
watershed sources, consistent with watershed uses, and consistent with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\citep{russell_effect_2006, caffreyHighNutrientPulses2007, peierlsNonmonotonicResponsesPhytoplankton2012, palmerImpactsDroughtsLow2015, ciraPhytoplanktonDynamicsLowInflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow were observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. Tidal flushing
from Matagorda Bay might be responsible for decreasing TKN
concentrations and limiting the effects of freshwater inflow in lower
reaches of Lavaca Bay. \citet{russell_effect_2006} suggested the
processing of organic loads in the upper portions of Lavaca Bay reduces
the transport of nutrients into the lower reaches of the Bay.

Freshwater inflow also had a strong positive effect on
chlorophyll-\emph{a} at the upper- and mid-bay sites. The upper-bay
site, TCEQ-13563, showed decreases in chlorophyll-\emph{a} at the
highest freshwater inflow volumes. Freshwater flushing or increases in
turbidity are associated with decreases in chlorophyll-\emph{a} in other
estuaries
\citep{peierlsNonmonotonicResponsesPhytoplankton2012, cloernPhytoplanktonPrimaryProduction2014}.
No relationships between inorganic nitrogen or TP loadings with
chlorophyll-\emph{a} were observed. Due to the lack of TKN loading
information, no assessment between organic nitrogen loads and
chlorophyll-\emph{a} were possible.

Although other studies have identified complex relationships between
estuary nutrient concentrations, nutrient loading and
chlorophyll-\emph{a} concentrations in Texas estuaries
\citep{ornolfsdottirNutrientPulsingRegulator2004, doradoUnderstandingInteractionsFreshwater2015, ciraPhytoplanktonDynamicsLowInflow2021, tominackVariabilityPhytoplanktonBiomass2022},
this study specifically used flow-adjusted freshwater derived nutrient
loads to parse out contributions from changes in nutrient loadings while
accounting for variations in load due to flow. Loading GAMs indicated no
evidence of changes in flow-normalized TP loads in either river, and no
changes in flow-normalized NO\textsubscript{3} loads in the Navidad
River. The small changes in flow-normalized NO\textsubscript{3} loads in
the Lavaca River are probably masked under most conditions by discharge
from the Navidad River. Given the relatively little variation in
flow-normalized loads, it can be expected that they would contribute
little to the variance in downstream water quality.

GAMs did not identify responses in DO concentration to inflows or
nutrient loads. The seasonality term in the temporal GAM models
explained a substantial amount of DO variation at all of the sites.
Responses of estuary metabolic processes and resulting DO concentrations
can be quite complicated and often locally specific
\citep{caffreyFactorsControllingNet2004}. While the lack of total
nitrogen or TKN loading data hinders interpretation, the large seasonal
effect on DO suggests physical factors play an important role and should
be included in future models. Prior work suggests that Lavaca Bay may
not be limited by nutrients alone, with high turbidity or nutrient
processing in upper portions of the Bay or intertidal river limiting
production \citep{russell_effect_2006}.

@@

GAM models appear to provide reliable estimates of nutrient loads in the
Lavaca Bay watershed. However, additional flow-biased data collection
efforts would decrease the prediction variance and probably improve
accuracy at critical high flow events. Ongoing projects will fill data
gaps for total nitrogen and TKN loading. This study, consistent with
others along the Texas coast, found strong effects of freshwater flow on
nutrient and chlorophyll-\emph{a} concentrations. DO concentrations,
dominated by seasonal effects, did not show strong direct responses to
freshwater flow. Small variance in flow-adjusted nutrient loads
indicates that (1) there have been limited changes in non point sources
of nutrients and (2) that those changes have limited effect on water
quality in Lavaca Bay to date. Although the study did not identify
strong responses to changes in nutrient loading, this does provide a
baseline assessment for future water quality management activities in
the watershed.

>>>>

This study used an information theoretic approach to evaluate if
nutrient loads and/or freshwater inflows provided evidence of effects on
water quality concentrations in Lavaca Bay. Model probabilities were
calculated and compared using the AIC\textsubscript{c} scores between
each group of temporal, flow, and flow+load models
\citep{burnhamAICModelSelection2011}. Improvements in model
probabilities provide evidence that the terms explain additional
variation in the response variable. If model probabilities were tied,
there wasn't evidence the more complicated model explains additional
variation in water quality.

@@

Using evaluation criteria from
\citet{moriasiHydrologicWaterQuality2015}, predictive performance of
nutrient loading GAMs ranged from ``satisfactory'' to ``very good''
based on median NSE, r\textsuperscript{2}, and PBIAS metrics calculated
using 5-fold cross validation. Median NSE values were 0.34 (Lavaca
River) and 0.48 (Navidad River) for NO\textsubscript{3} loads and 0.81
(Lavaca River) and 0.91 (Navidad River) for TP loads. Median
r\textsuperscript{2} values were 0.70 (Lavaca River) and 0.87 (Navidad
River) for NO\textsubscript{3} loads and 0.93 (Lavaca River) and 0.99
(Navidad River) fopr TP loads. Median PBIAS values were 2.00 (Lavaca
River) and 10.90 (Navidad River) for NO\textsubscript{3} loads and -7.20
(Lavaca River) and -3.30 (Navidad River) for TP loads. Density plots of
metrics show similar distribution of values between sites for the same
parameter, with the exception r\textsuperscript{2} values for
NO\textsubscript{3} loads where USGS-08164000 showed a much larger
variance in values compared to USGS-08164525 (Figure \ref{fig:fig2}). TP
GAMS had higher average NSE and r\textsuperscript{2} values and lower
variance in metric values compared to NO\textsubscript{3}.

@@

Predicted annual NO\textsubscript{3} and TP loads show considerable
variation, generally following patterns in discharge (Figure
\ref{fig:fig3}, Figure \ref{fig:fig4}). Flow-normalized TP loads at both
sites and flow-normalized NO\textsubscript{3} loads at USGS-08164000
indicate watershed based loads did not change much over time when
accounting for variation driven by streamflow (Figure \ref{fig:fig3}).
Flow-normalized loads at USGS-08164000 show small variation over time
with some decreases in NO\textsubscript{3} loads since 2013.

Aggregated across both sites, the mean annual NO\textsubscript{3} load
2005 through 2020 was 205,405 kg (126,867 kg - 341,569 kg, 90\% CI).
Annual NO\textsubscript{3} loads ranged from 12,574 kg in 2011 to
794,510 kg in 2007. Total annual TP loads ranged from 7,839 kg in 2011
to 595,075 kg in 2007. Mean annual TP loading from 2005 through 2020 was
182,673 kg (152,227 kg - 219,310 kg, 90\% CI). On average, USGS-08164525
accounted for 68\% of NO\textsubscript{3} loads and 59\% of TP loads
from 2005 through 2020. However, during periods of extreme drought the
Lavaca River (USGS-08164000) became the primary source of nutrient
loading in the watershed with the Navidad River only accounting for 15\%
and 25\% of NO\textsubscript{3} and TP loads in 2011 (Figure
\ref{fig:fig4}).

@@

Converted to average annual yield, the estimates of annual TP loads at
Lavaca River (USGS-08164000) are within the ranges in previous published
studies (Table \ref{tab:tpcomp};
\citep{dunnTrendsNutrientInflows1996, rebich_sources_2011, omaniEstimationSedimentNutrient2014, wise_spatially_2019}).
It isn't obvious why TP estimates in
\citet{dunnTrendsNutrientInflows1996} were notably lower. Given that
none of the studies identify substantially sized trends in TP, it is is
possible that the period used in \citet{dunnTrendsNutrientInflows1996}
was drier on average than the other studies. The SPARROW models used in
\citet{rebich_sources_2011} and \citet{wise_spatially_2019} utilize a
version of LOADEST in the underlying load estimation procedure, so a
difference due to methodology alone is unlikely.

@@

Cross-validation of the GAM loading models indicated that GAMs performed
well on average at predicting daily nutrient loading values. The
variance in scores was very high indicating subsets of values were
problematic at characterizing functional relationships between nutrients
and predictors. Because all of the water quality data for these two
locations in the TCEQ databases were ambient water quality data,
collected to be representative of typical flow conditions, there were
few data at the highest portions of the flow-duration curve. It was
beyond the scope of the current study to evaluate the subsets of
cross-validation data and scores. However, the cross-validation
procedure is indicative that more robust sampling would be beneficial
for reducing prediction variance. Supplementary flow-biased monitoring
targeting storm- or high-flow conditions is recommended here to improve
the precision of GAM predictions
\citep{horowitzEvaluationSedimentRating2003, snelderEstimationCatchmentNutrient2017}.

The non-linear temporal water quality trends identified using GAMs
differed slightly from the trends identified by
\citet{bugica_water_2020}. This is not unexpected due to the different
time periods, different methodology, and generally small slopes
identified for most of the significant water quality parameters in prior
work. The trend in DO and cholorophyll-\emph{a} concentrations are
stable in comparison to other Texas estuaries that are facing larger
demands for freshwater diversions, higher population growth, and more
intense agricultural production
\citep{wetzWaterQualityDynamics2016, bugica_water_2020}. The trend of
increasing NO\textsubscript{\emph{x}} concentration at the upper-bay
TCEQ-13563 site and recent increases in TKN concentration at the mid-bay
TCEQ-13383 site are concerning due to the nitrogen limitation identified
in many Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the relatively low ambient concentrations observed in Lavaca Bay.

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{x}}, TKN, and TP are suggestive of nonpoint
watershed sources, consistent with watershed uses and with other studies
relating freshwater inflow with nutrient concentrations in Lavaca Bay
and other estuaries
\citep{russell_effect_2006, caffreyHighNutrientPulses2007, peierlsNonmonotonicResponsesPhytoplankton2012, palmerImpactsDroughtsLow2015, ciraPhytoplanktonDynamicsLowInflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow were observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. Tidal flushing
from Matagorda Bay could be responsible for diluting TKN and acting as a
control on the effects of freshwater inflow in lower reaches of Lavaca
Bay. \citet{russell_effect_2006} also suggested the processing of
organic loads in the upper portions of Lavaca Bay reduces the transport
of nutrients into the lower reaches of the Bay.

Freshwater inflow had a strong positive effect on chlorophyll-\emph{a}
at the upper- and mid-bay sites. The upper-bay site, TCEQ-13563, showed
decreases in chlorophyll-\emph{a} at the highest freshwater inflow
volumes. Freshwater flushing or increases in turbidity are associated
with decreases in chlorophyll-\emph{a} in other estuaries
\citep{peierlsNonmonotonicResponsesPhytoplankton2012, cloernPhytoplanktonPrimaryProduction2014}.
No relationships between inorganic nitrogen or TP loadings with
chlorophyll-\emph{a} were observed. Due to the lack of TKN loading
information, no assessment between organic nitrogen loads and
chlorophyll-\emph{a} were possible.

Although other studies have identified complex relationships between
estuary nutrient concentrations, nutrient loading and
chlorophyll-\emph{a} concentrations in Texas estuaries
\citep{ornolfsdottirNutrientPulsingRegulator2004, doradoUnderstandingInteractionsFreshwater2015, ciraPhytoplanktonDynamicsLowInflow2021, tominackVariabilityPhytoplanktonBiomass2022},
this study specifically used flow-adjusted freshwater derived nutrient
loads to parse out contributions from changes in nutrient loadings while
accounting for variations in load due to flow. Loading GAMs indicated no
evidence of changes in flow-normalized TP loads in either river, and no
changes in flow-normalized NO\textsubscript{3} loads in the Navidad
River. The small changes in flow-normalized NO\textsubscript{3} loads in
the Lavaca River are probably masked under most conditions by discharge
from the Navidad River. Given the relatively small variation in
flow-normalized loads, it can be expected that they would contribute
little to the variance in downstream water quality.

GAMs did not identify responses in DO concentration to inflows or
nutrient loads. The seasonality term in the temporal GAM models
explained a substantial amount of DO variation at all of the sites.
Responses of estuary metabolic processes and resulting DO concentrations
can be quite complicated and often locally specific
\citep{caffreyFactorsControllingNet2004}. While the lack of total
nitrogen or TKN loading data hinders interpretation, the large seasonal
effect on DO suggests physical factors play an important role and should
be included in future models. Prior work suggests that Lavaca Bay may
not be limited by nutrients alone, with high turbidity or nutrient
processing in upper portions of the Bay or intertidal river limiting
production \citep{russell_effect_2006}. Finally, it is reasonable to
assume that fluctuations in DO may not occur immediately in response to
nutrient pulses or freshwater inflow. Work has has shown that various
water quality parameters may have lagged effects lasting days or even
months following storms and large discharge events
\citep{mooney_watershed_2012, wetzExtremeFutureEstuaries2013, bukaveckas_influence_2020, walkerTimescalesMagnitudeWater2021}.
However, our work only evaluates responses to loading and inflows
occuring the day of water quality observations.

@@

GAM models appear to provide reliable estimates of nutrient loads in the
Lavaca Bay watershed. However, additional flow-biased data collection
efforts would decrease the prediction variance and improve accuracy at
critical high flow events. Ongoing projects will fill data gaps for
total nitrogen and TKN loading. This study, consistent with others along
the Texas coast, found strong effects of freshwater flow on nutrient and
chlorophyll-\emph{a} concentrations. DO concentrations, dominated by
seasonal effects, did not show strong direct responses to freshwater
flow. Small variance in flow-adjusted nutrient loads indicates that (1)
there have been limited changes in non-point sources of nutrients and
(2) that there isn't strong evidence that those small changes have had
effects on chlorophyll-\emph{a} or dissolved oxygen in Lavaca Bay.
Although the study did not identify strong responses to changes in
nutrient loading, this does provide a baseline assessment for future
water quality management activities in the watershed.
